Vectorizing Python Constructs Using Pythran and Boost SIMD

### 2013-11-27 doc/papers/wpmvp14/paper.tex
First Pierrick review

The Python language has grown in audience for the past ten years, even reaching
the world of scientific computations thanks to the \texttt{numpy} module, a
module that provides a MATLAB-like API. As a consequence, more and more code is
being written either in pure Python, generally to prototype an application, or
as a Python and native code mix when performance matters. However, Python
trades performance for dynamicity and does not particularly shines in terms of
performance.

@@

A native module is a shared library whose interface make it callable from the
Python interpreter using the regular \texttt{import} mechanism. It makes it
possible to call FORTRAN/C/C++ code from Python in a transparent way. Typically
Python object are unboxed when transferred from Python to C, where efficient
computations can happen. The the result is boxed when getting back to the
Python world. In between, vectorization can happen.  This is the approach taken
by the \texttt{numpy} module: a native type, the \texttt{ndarray} contains
contiguous unboxed elemnts, and provides a great deal of high-level functions
to manipulate them. All the computation-intensive part is done in pure C.
Vectorization can eventually happen at this level, for instance when performing
a point-to-point operation like the sum of two arrays.  However the granuarity
of the operations limits the ratio of load / stores versus operation. For
instance, if \texttt{a}, \texttt{b} and \texttt{c} are 1-D arrays, the sum
\texttt{a + b +c} is computed using two loops and an intermediate array.

AOT compilation uses type inference or type annotation to statically compute
the type of all (like in Shedskin, Parakeet) or part of (like in Cython or
Numba) variables, effectively unboxing them to speed-up computations. Using JIT
compilation, type information is available at runtime and the unboxing decision
is based on profitablity, e.g.\ according to a trace analysis. This is the
approach of the PyPy project. In both cases, generation of vector instruction
is possible, but none of the cited project use it.  Indeed automatic
vectorization is a complex topic, and as stated in~\cite{maleki2011}, many cases
are not yet correctly supported even for mainstream compilers.

Pythran is also an AOT compiler, which holds the unique property of turning
Python modules into C++ \emph{meta-programs}. It does not need type annotations
and keeps the polymorphism of original functions, respecting Python's duck
typing. It compilation frow is described in Figure~\ref{fig:pythran-compiler}.
Like other AOT compilers, it unboxes all variables for better performance, and
performs a wide variety of optimizations such as lazy evaluation, constant
folding, expression fusion etc.  It supports core \texttt{numpy} constructs and
most Python constructs to the notable exception of user classes. Constructs
such as \texttt{eval}, are not supported.

@@

\subsection{Load/Store Protocol}

@@

A high-level language such as Python exhibit another level of vectorization
opportunity in implict loop. Implicit loops can be foud in many places, but
this paper focuses on \emph{list comprehension} and \emph{generator
expression}, \emph{map} or \emph{sum} intrinsics and \emph{array operations}
through the \texttt{numpy} module.

@@

\subsubsection{\texttt{sum}}

The \texttt{sum} intrinsic is a typical case of potential vectorization of an
implict loop. Its prototype is: \texttt{sum(sequence[, start]) $\rightarrow$
value} and its behavior is equivalent to the Python code in
Listing~\ref{lst:sum-py}. Depending on the type of \texttt{sequence}, and also
wether one needs to strictly conforms to the IEEE Standard for Floating-Point
Arithmetic (IEEE 754).

@@

\begin{enumerate}
    \item\label{enu:same-types} All element of \texttt{sequence} have the same types;
    \item\label{enu:scalar-types} \texttt{sequence} contains scalar types;
    \item\label{enu:contiguous} \texttt{sequence} contains contiguous elements;
    \item\label{enu:random-access} \texttt{sequence} provides a random access iterator. in particular one can computes its size in $\mathcal{O}(1)$.
\end{enumerate}

Condition~\ref{enu:same-types} is a prerequiste for static typing and efficient
unboxing. For instance, any code that does not match it will fail to compile
using Pythran or Shed-Skin. It is also satisified for the \texttt{ndarray}
container as they wraps unboxed arrays. Condition~\ref{enu:scalar-types} can be
checked at compile time. Python supports two native types candidates for
vectorization: \texttt{int} (64 bits integer) and \texttt{float}
(double-precision floating point numbers). The \texttt{numpy} model supports a
wide range of signed and unsigned integers as well as single, double or
quadruple precision float. Depending on the target architecture, only some of
them can fit into vector registers. Finally condition~\ref{enu:contiguous} also
is a static property of a container. It is verified for C++ alternative to
Python's \texttt{list}, \texttt{str} and \texttt{ndarrays}, but generally not
for \texttt{set} or \texttt{dict}. Condition \ref{enu:random-access} is not
matched by \emph{generator}s, the Python incarnation of co-routines.

Depending on the container properties, it is possible to select a vectorized
implementation of fallback to a generic one, e.g.\ using Substitution Failure
Is Not An Error (SFINAE).

@@

Generator expression is somehow similar to list comprehension, to the notable
exception that the value returned is an \emph{iterator adaptator} on the
container argument. In the context of vectorization, a generator expression
supports the vector protocol if the associated container also supports it. In
that case the  \emph{iterator adaptator} implements the vector protocol by
applying the comprehension rule to each vctor element from this container, in a
streaming fashion.

>>>>

The Python language has grown in audience for the past ten years, even reaching
the world of scientific computations thanks to the \texttt{numpy} module, a
%% what about the scipy module? It is not really used?
module that provides a MATLAB-like API. As a consequence, more and more code is
being written either in pure Python, generally to prototype an application, or
as a Python and native code mix when performance matters. However, Python
trades performance for dynamicity and does not particularly shines in terms of
performance.

@@

A native module is a shared library whose interface make it callable from the
Python interpreter using the regular \texttt{import} mechanism. It makes it
possible to call native code from Python in a transparent way. Typically
%% It may be others languages, isn't it?
Python object are unboxed when transferred from Python to C, where efficient
computations can happen. Then the result is boxed when getting back to the
Python world. In between, vectorization can happen. This is the approach taken
by the \texttt{numpy} module: a native type, the \texttt{ndarray} contains
contiguous unboxed elements, and provides a great deal of high-level functions
to manipulate them. All the computation-intensive part is done in pure C.
Vectorization can eventually happen at this level, for instance when performing
a point-to-point operation like the sum of two arrays. However the granularity
of the operations limits the ratio of \texttt{load}/\texttt{store} versus operation. For
instance, if \texttt{a}, \texttt{b} and \texttt{c} are 1-D arrays, the sum
\texttt{a + b + c} is computed using two loops and an intermediate array.
% It means there are 4 load / stores and only two loops of addition.

AOT compilation uses type inference or type annotation to statically compute
the type of all (like in Shedskin, Parakeet) or part of (like in Cython or
Numba) variables, effectively unboxing them to speed-up computations. Using JIT
compilation, type information is available at runtime and the unboxing decision
is based on profitability, e.g.\ according to a trace analysis. This is the
approach of the PyPy project. In both cases, generation of vector instruction
is possible, but none of the cited project use it. Indeed automatic
vectorization is a complex topic, and as stated in~\cite{maleki2011}, many cases
are not yet correctly supported even for mainstream compilers.

Pythran is also an AOT compiler, which holds the unique property of turning
Python modules into C++ \emph{meta-programs}. It does not need type annotations
and keeps the polymorphism of original functions, respecting Python's duck
typing. Its compilation workflow is described in Figure~\ref{fig:pythran-compiler}.
Like others AOT compilers, it unboxes all variables for better performance, and
performs a wide variety of optimizations such as lazy evaluation, constant
folding, expression fusion etc. It supports core \texttt{numpy} constructs and
most Python constructs to the notable exception of user classes. Constructs
such as \texttt{eval}, are not supported.

@@

\subsection{\texttt{Load}/\texttt{Store} Protocol}

@@

A high-level language such as Python exhibit another level of vectorization
opportunity in implict loop. Implicit loops can be found in many places, but
this paper focuses on \emph{list comprehension} and \emph{generator
expression}, \emph{map} or \emph{sum} intrinsics and \emph{array operations}
through the \texttt{numpy} module.

@@

\subsubsection{\texttt{sum}}
% only sum as texttt in a title is not really beautiful

The \texttt{sum} intrinsic is a typical case of potential vectorization of an
implict loop. Its prototype is: \texttt{sum(sequence[, start]) $\rightarrow$
value} and its behavior is equivalent to the Python code in
Listing~\ref{lst:sum-py}. Depending on the type of \texttt{sequence}, and also
wether one needs to strictly conforms to the IEEE Standard for Floating-Point
Arithmetic (IEEE 754).
% I don't understand what this last sentence is doing here :-s

@@

\begin{enumerate}
    \item\label{enu:same-types} All element of \texttt{sequence} have the same types;
    \item\label{enu:scalar-types} \texttt{sequence} contains scalar types;
    \item\label{enu:contiguous} \texttt{sequence} contains contiguous elements;
    \item\label{enu:random-access} \texttt{sequence} provides a random access iterator. In particular one can computes its size in $\mathcal{O}(1)$.
\end{enumerate}

Condition~\ref{enu:same-types} is a prerequiste for static typing and efficient
unboxing. For instance, any code that does not match it will fail to compile
using Pythran or Shed-Skin. It is also satisified for the \texttt{ndarray}
container as they wraps unboxed arrays. Condition~\ref{enu:scalar-types} can be
checked at compile time. Python supports two native scalar types candidates for
vectorization: \texttt{int} (64 bits integer) and \texttt{float}
(double-precision floating point numbers). The \texttt{numpy} model add supports for a
wide range of signed and unsigned integers as well as single, double or
quadruple precision float. Depending on the target architecture, only some of
them can fit into vector registers. Finally condition~\ref{enu:contiguous} also
is a static property of a container. It is verified for C++ alternative to
Python's \texttt{list}, \texttt{str} and \texttt{ndarrays}, but generally not
for \texttt{set} or \texttt{dict}. Condition \ref{enu:random-access} is not
matched by \emph{generator}s, the Python incarnation of co-routines.

%we know a work have to be done about random acces but you speak about something else now ...
%It is disturbing and we don't really know why you speak about this new topic.

As function have to be generic but some containers do not match requierement
properties, it is possible to select a vectorized
implementation of fallback to a generic one using Substitution Failure
Is Not An Error (SFINAE).

@@

Generator expression is somehow similar to list comprehension, to the notable
exception that the value returned is an \emph{iterator adaptator} on the
container argument. In the context of vectorization, a generator expression
supports the vector protocol if the associated container also supports it. In
that case the  \emph{iterator adaptator} implements the vector protocol by
applying the comprehension rule to each vector element from this container, in a
streaming fashion.

Python also provide same features for set and dictionnary.
